---
title: "GIS Demonstration"
author: "Tim Norris"
date: "2024-02-20"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GIS Demonstration

GIS Demonstration.R 2024, February Tim Norris
[tnorris\@miami.edu](mailto:tnorris@miami.edu){.email}

------------------------------------------------------------------------

This is free software: you can redistribute it and/or modify it under
the terms of the GNU General Public License as published by the Free
Software Foundation, either version 3 of the License, or any later
version.

This code is distributed in the hope that it will be useful, but WITHOUT
ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
FITNESS FOR A PARTICULAR PURPOSE. See the GNU General Public License for
more details. <https://www.gnu.org/licenses/>.

------------------------------------------------------------------------

### Load the needed libraries

```{r}
# https://leafletjs.com
library(leaflet)
# https://r-spatial.github.io/sf/articles/sf1.html
library(sf)
library(jqr)
library(curl)
library(ggplot2)
library(dplyr)
library(spdep)
library(spatialreg)
library(spgwr)
```

### Get the Data

from: <https://gdsc.idsc.miami.edu> explore the metadata at:
<https://gdsc.idsc.miami.edu/detail/mdc_2021_acs_5yr_tract_dvmt>

```{r}
base_api_url <- "https://gdsc.idsc.miami.edu/functions/"
table_name <- "mdc_2021_acs_5yr_tract_dvmt_geojson"
full_api_url <- paste0(base_api_url,table_name)
```

| Paste the output from the next line into an empty web browser tab and review
| the data. -- Try to find the EPSG code -- What are the feature attributes in
| this data?

```{r}
print(full_api_url)
```

#### Read the Data with Curl

Note that EPSG 4269 is the native coordinate reference system:\
NAD83 lat/lng

```{r include=FALSE}
# include=FALSE suppresses output
tracts_json <- curl::curl(full_api_url)
tracts_geojson <- jqr::jq(tracts_json,'.[]["data"]')
tracts <- sf::st_read(tracts_geojson,crs='EPSG:4269')
```

#### Re-project the Data into Needed Coordinate Reference Systems

EPSG 2236: NAD83 State Plane Florida East Feet (for measured
distances).\
EPSG 3857: WGS84 Web Mercator (for visualizing in RStudio on the
internet).

```{r}
tracts_sp <- sf::st_transform(tracts,crs='EPSG:2236') 
tracts <- sf::st_transform(tracts,crs='EPSG:4326') # WGS84 lat/lng
```

#### cleanup

```{r}
rm(tracts_geojson)
rm(tracts_json)
```

### Explore the Data

```{r}
class(tracts)
attr(tracts,"sf_column")
```

| Note the geometry column in the output from the next line

```{r}
print(tracts[c(1,2,3,4,15)], n=3)
```

| Take a moment to really look at the list from the next line - lots of data
| for telling stories here ...

```{r}
names(tracts)
```

```{r}
summary(tracts$median_income_of_residents)
```

| Note the no data value of -666666666, this 'nodata' number from the US Census
| Bureau will cause problems. Summarize again ignoring these values ...

```{r}
tracts$median_income_of_residents <- na_if(
  tracts$median_income_of_residents,-666666666
)
summary(tracts$median_income_of_residents)
```

| Note the NA's (what do they mean? how can we control for this?)

#### Summarize some Other Columns

```{r}
summary(tracts$total_population)
```

| Note that some variables need to be normalized (why??)
| Normalize by total population.

```{r}
summary(tracts$white_alone)
summary(tracts$white_alone/tracts$total_population)
```

| Normalize by total_housing_units

```{r}
summary(tracts$households_below_poverty_level)
summary(tracts$total_housing_units)
summary(tracts$households_below_poverty_level/tracts$total_housing_units)
```

| Note the NA's (what do they mean? how can we control for this?)
| Also note the Max is greater than 1!! For a normalized value this cannot be!!
| There must be an error either in how it was normalized or in the actual counts.

### Make an Income Map (Choropleth)

First set some color bins for the data.\
Note first 'palette' parameter comes from <https://colorbrewer2.org>.\
You can experiment with different palettes.

```{r}
median_income_pal <- colorBin(
  palette = "YlGnBu",                        # color ramp
  tracts$median_income_of_residents,         # data column
  7                                          # number of bins
)
```

Set output dimension

```{r}
output.height <- '800px'
```

Create map object

```{r}
m_median_income <- leaflet(height=800) %>%
  addTiles() %>% # Add default OpenStreetMap map tiles
  addPolygons(
    data = tracts,
    fillColor = ~median_income_pal(median_income_of_residents),
    fillOpacity = 0.8,
    weight = 0.3,
    label = ~paste('median income in US$: ',median_income_of_residents),
    labelOptions = labelOptions(style = list(
      "font-weight" = "normal",
      padding = "3px 8px"),
      textsize = "13px", direction = "auto"),
    highlight = highlightOptions(color = "black", weight = 4, bringToFront = TRUE))

# draw the map
m_median_income
```

| Take a moment to explore. Note the mouse-over value, what is it (label)?
| -- edit some of the parameters and rerun the code to see changes
| -- find the needed changes to add a popup on click

### Make a Race Map (Choropleth)

Make a normalized column (percentages better for maps).

```{r}
tracts$percent_white <- tracts$white_alone/tracts$total_population
```

First set some color bins for the data.

```{r}
percent_white_pal <- colorBin(
  "OrRd", 
  tracts$percent_white, 
  7)
```

Set output dimension

```{r}
output.height <- '800px'
```

Create map object.

```{r}
m_percent_white <- leaflet(height=800) %>%
  addTiles() %>% # Add default OpenStreetMap map tiles
  addPolygons(
    data = tracts,
    fillColor = ~percent_white_pal(tracts$percent_white),
    fillOpacity = 0.8,
    weight = 0.3,
    label = ~paste('percent white: ',tracts$percent_white),
    labelOptions = labelOptions(style = list(
      "font-weight" = "normal",
      padding = "3px 8px"),
      textsize = "13px", direction = "auto"),
    popup = ~paste('<strong>white_alone:',white_alone,'</strong><br>Total Pop:',total_population),
    highlight = highlightOptions(color = "black", weight = 4, bringToFront = TRUE))

# draw the map
m_percent_white
```

| Take a moment to explore, does this match your expectations?
| Is there anything that does or does not surprise you?

### Basic Linear Model: median_income \~ percent_white

We will model median income as a function of percent white.\
What do you expect this model to look like? Which way will the\
correlation be sloped?

Make a linear regression model and look at the summary.

```{r}
nonspatial <- lm(percent_white~median_income_of_residents,data=tracts)
summary(nonspatial)
```

It looks like we have a strong relationship!! The p-values for the
intercept and percent_with predictor are very low.

Make a plot of the basic linear model.

```{r}
options(repr.plot.width = 8, repr.plot.height = 8)
plot(tracts$percent_white,tracts$median_income_of_residents)
abline(nonspatial, col="red")
```

| Take a moment to check-in - does this meet your expectations or hypothesis?

Make a basic q-q plot.

```{r}
qqnorm(nonspatial$residuals)
```

| For a q-q plot if the line is not perfectly diagonal, we may be breaking
| some basic assumptions about linear models; specifically our data does
| not have a normal distribution.

Calculate the residuals

```{r}
x = data.frame(
  actual = nonspatial$fitted.values + nonspatial$residuals,
  fitted = nonspatial$fitted.values)
x = x[order(x$fitted),]
```

Make a quick plot of the residuals

```{r}
plot(x$actual, col='#00000040', ylab="percent white", las=1)
lines(x$fitted, col='#c00000', lwd=3)
```

#### INTERPRETATION

The model suggests a relationship between percent white and median
income exists,\
BUT The R-squared suggests only 15% of explanation, seems that there is
a lot\
of other stuff going on as well. This lack of explanation is also shown
in the\
residual plot directly above (not tightly clustered). The Q-Q plot
suggests a\
distribution that is skewed to the right.

### 
