---
title: "GIS Demonstration"
author: "Tim Norris"
date: "2024-02-20"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GIS Demonstration

GIS Demonstration.R 2024, February Tim Norris
[tnorris\@miami.edu](mailto:tnorris@miami.edu){.email}

------------------------------------------------------------------------

This is free software: you can redistribute it and/or modify it under
the terms of the GNU General Public License as published by the Free
Software Foundation, either version 3 of the License, or any later
version.

This code is distributed in the hope that it will be useful, but WITHOUT
ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
FITNESS FOR A PARTICULAR PURPOSE. See the GNU General Public License for
more details. <https://www.gnu.org/licenses/>.

------------------------------------------------------------------------

### Load the needed libraries

```{r}
# https://leafletjs.com
library(leaflet)
# https://r-spatial.github.io/sf/articles/sf1.html
library(sf)
library(jqr)
library(curl)
library(ggplot2)
library(dplyr)
library(spdep)
library(spatialreg)
library(spgwr)
```

### Get the Data

from: <https://gdsc.idsc.miami.edu> explore the metadata at:
<https://gdsc.idsc.miami.edu/detail/mdc_2021_acs_5yr_tract_dvmt>

```{r}
base_api_url <- "https://gdsc.idsc.miami.edu/functions/"
table_name <- "mdc_2021_acs_5yr_tract_dvmt_geojson"
full_api_url <- paste0(base_api_url,table_name)
```

| Paste the output from the next line into an empty web browser tab and review
| the data. -- Try to find the EPSG code -- What are the feature attributes in
| this data?

```{r}
print(full_api_url)
```

#### Read the Data with Curl

Note that EPSG 4269 is the native coordinate reference system:\
NAD83 lat/lng

```{r include=FALSE}
# include=FALSE suppresses output
tracts_json <- curl::curl(full_api_url)
tracts_geojson <- jqr::jq(tracts_json,'.[]["data"]')
tracts <- sf::st_read(tracts_geojson,crs='EPSG:4269')
```

#### Re-project the Data into Needed Coordinate Reference Systems

EPSG 2236: NAD83 State Plane Florida East Feet (for measured
distances).\
EPSG 3857: WGS84 Web Mercator (for visualizing in RStudio on the
internet).

```{r}
tracts_sp <- sf::st_transform(tracts,crs='EPSG:2236') 
tracts <- sf::st_transform(tracts,crs='EPSG:4326') # WGS84 lat/lng
```

#### cleanup

```{r}
rm(tracts_geojson)
rm(tracts_json)
```

### Explore the Data

```{r}
class(tracts)
attr(tracts,"sf_column")
```

| Note the geometry column in the output from the next line

```{r}
print(tracts[c(1,2,3,4,15)], n=3)
```

| Take a moment to really look at the list from the next line - lots of data
| for telling stories here ...

```{r}
names(tracts)
```

```{r}
summary(tracts$median_income_of_residents)
```

| Note the no data value of -666666666, this 'nodata' number from the US Census
| Bureau will cause problems. Summarize again ignoring these values ...

```{r}
tracts$median_income_of_residents <- na_if(
  tracts$median_income_of_residents,-666666666
)
summary(tracts$median_income_of_residents)
```

| Note the NA's (what do they mean? how can we control for this?)

#### Summarize some Other Columns

```{r}
summary(tracts$total_population)
```

| Note that some variables need to be normalized (why??)
| Normalize by total population.

```{r}
summary(tracts$white_alone)
summary(tracts$white_alone/tracts$total_population)
```

| Normalize by total_housing_units

```{r}
summary(tracts$households_below_poverty_level)
summary(tracts$total_housing_units)
summary(tracts$households_below_poverty_level/tracts$total_housing_units)
```

| Note the NA's (what do they mean? how can we control for this?)
| Also note the Max is greater than 1!! For a normalized value this cannot be!!
| There must be an error either in how it was normalized or in the actual counts.

### Make an Income Map (Choropleth)

First set some color bins for the data.\
Note first 'palette' parameter comes from <https://colorbrewer2.org>.\
You can experiment with different palettes.

```{r}
median_income_pal <- colorBin(
  palette = "YlGnBu",                        # color ramp
  tracts$median_income_of_residents,         # data column
  7                                          # number of bins
)
```

Set output dimension

```{r}
output.height <- '800px'
```

Create map object

```{r}
m_median_income <- leaflet(height=800) %>%
  addTiles() %>% # Add default OpenStreetMap map tiles
  addPolygons(
    data = tracts,
    fillColor = ~median_income_pal(median_income_of_residents),
    fillOpacity = 0.8,
    weight = 0.3,
    label = ~paste('median income in US$: ',median_income_of_residents),
    labelOptions = labelOptions(style = list(
      "font-weight" = "normal",
      padding = "3px 8px"),
      textsize = "13px", direction = "auto"),
    highlight = highlightOptions(color = "black", weight = 4, bringToFront = TRUE))

# draw the map
m_median_income
```

| Take a moment to explore. Note the mouse-over value, what is it (label)?
| -- edit some of the parameters and rerun the code to see changes
| -- find the needed changes to add a popup on click

### Make a Race Map (Choropleth)

Make a normalized column (percentages better for maps).

```{r}
tracts$percent_white <- tracts$white_alone/tracts$total_population
```

First set some color bins for the data.

```{r}
percent_white_pal <- colorBin(
  "OrRd", 
  tracts$percent_white, 
  7)
```

Set output dimension

```{r}
output.height <- '800px'
```

Create map object.

```{r}
m_percent_white <- leaflet(height=800) %>%
  addTiles() %>% # Add default OpenStreetMap map tiles
  addPolygons(
    data = tracts,
    fillColor = ~percent_white_pal(tracts$percent_white),
    fillOpacity = 0.8,
    weight = 0.3,
    label = ~paste('percent white: ',tracts$percent_white),
    labelOptions = labelOptions(style = list(
      "font-weight" = "normal",
      padding = "3px 8px"),
      textsize = "13px", direction = "auto"),
    popup = ~paste('<strong>white_alone:',white_alone,'</strong><br>Total Pop:',total_population),
    highlight = highlightOptions(color = "black", weight = 4, bringToFront = TRUE))

# draw the map
m_percent_white
```

| Take a moment to explore, does this match your expectations?
| Is there anything that does or does not surprise you?

### Basic Linear Model: median_income \~ percent_white

We will model median income as a function of percent white.\
What do you expect this model to look like? Which way will the\
correlation be sloped?

Make a linear regression model and look at the summary.

```{r}
nonspatial <- lm(percent_white~median_income_of_residents,data=tracts)
summary(nonspatial)
```

It looks like we have a strong relationship!! The p-values for the
intercept and percent_with predictor are very low.

Make a plot of the basic linear model.

```{r}
options(repr.plot.width = 8, repr.plot.height = 8)
plot(tracts$percent_white,tracts$median_income_of_residents)
abline(nonspatial, col="red")
```

| Take a moment to check-in - does this meet your expectations or hypothesis?

Make a basic q-q plot.

```{r}
qqnorm(nonspatial$residuals)
```

| For a q-q plot if the line is not perfectly diagonal, we may be breaking
| some basic assumptions about linear models; specifically our data does
| not have a normal distribution.

Calculate the residuals

```{r}
x = data.frame(
  actual = nonspatial$fitted.values + nonspatial$residuals,
  fitted = nonspatial$fitted.values)
x = x[order(x$fitted),]
```

Make a quick plot of the residuals

```{r}
plot(x$actual, col='#00000040', ylab="percent white", las=1)
lines(x$fitted, col='#c00000', lwd=3)
```

#### INTERPRETATION

The model suggests a relationship between percent white and median
income exists,\
BUT The R-squared suggests only 15% of explanation, seems that there is
a lot\
of other stuff going on as well. This lack of explanation is also shown
in the\
residual plot directly above (not tightly clustered). The Q-Q plot
suggests a\
distribution that is skewed to the right.

### Maps with GGPlot and Projected CRS

Median income

```{r}
plot(tracts_sp[34]) 
```

| Oops, remember the no data values?

```{r}
tracts_sp$median_income_of_residents <- na_if(
  tracts_sp$median_income_of_residents,-666666666
)
plot(tracts_sp[34])
```

Normalized percent white

```{r}
tracts_sp$percent_white <- tracts_sp$white_alone/tracts_sp$total_population
plot(tracts_sp[56])
```

### Check for spatial auto-correlation - Global Moran's I

There is a basic problem in all geo-spatial data analysis related to\
Tobler's First Law: everything is related to everything else, but near\
things are more related than distant things. This is reflected in the\
skewed Q-Q plot ... our independent variable is *not* independent, but\
but instead varies according to spatial proximity to other data points.\
This is known as spatial autocorrelation. The Moran's I statistic can\
confirm if we have spatial autocorrelation in our data.

<https://rspatial.org/raster/analysis/3-spauto.html>

$I = \frac{n}{\sum_{i=1}^n (y_i - \bar{y})^2} \frac{\sum_{i=1}^n \sum_{j=1}^n w_{ij}(y_i - \bar{y})(y_j - \bar{y})}{\sum_{i=1}^n \sum_{j=1}^n w_{ij}}$

First remove all NA values

```{r}
tracts_sp <- na.omit(tracts_sp)
```

Create a list of neighbors for each polygon.

```{r}
w <- poly2nb(tracts_sp, row.names=tracts_sp$fips)
class(w)
summary(w)
```

#### Show neighboring tracts.

```{r}
xy <- st_centroid(tracts_sp$geometry)
class(xy)
```

Transform neighbor object to a sf object. Note that we have to set the
CRS otherwise the coordinates are just numbers and not actually located
on the Earth's surface.

```{r}
w_sf <- as(nb2lines(w, coords = xy), 'sf')
w_sf <- st_set_crs(w_sf, st_crs(tracts_sp))
```

Make plot

```{r}
options(repr.plot.width = 16, repr.plot.height = 16)
ggplot() + 
  geom_sf(data=tracts_sp,aes(fill = percent_white)) + 
  scale_fill_binned(type = "viridis") +
  geom_sf(data=xy,color="white",size=0.5) + 
  geom_sf(data=w_sf,color="white",alpha=0.25) + 
  coord_sf(datum=NA)
```

Transform neighbors into spatial weights matrix

```{r}
wm <- nb2mat(w, style='B')
class(wm)
```

Get the matrix as a list

```{r}
ww <- nb2listw(w, style='B')
class(ww)
```

Calculate Moran's I (from spdep library)

```{r}
moran(tracts_sp$percent_white, ww, n=length(ww$neighbours), S0=Szero(ww))
```

#### INTERPRETATION Moran's I

Moran's I varies from -1 to 1. The further away from 0, the more
spatial\
autocorrelation - with a value of 0.68, we clearly show spatial\
autocorrelation.

```{r}
moran.test(nonspatial$residuals, ww, zero.policy=T)
moran.plot(nonspatial$residuals, ww, zero.policy=T)
```

The Moran's I statistic of 0.45 for the residuals, and the Moran plot of
the\
residuals showing a slight upward trend from left to right, confirms
this\
observation of spatial autocorrelation.

Instead of the approach above you should use Monte Carlo simulation.
That\
is the preferred method (in fact, the only good method). The way it
works\
that the values are randomly assigned to the polygons, and the Moran's
I\
is computed. This is repeated several times to establish a distribution
of\
expected values. The observed value of Moran's I is then compared with
the\
simulated distribution to see how likely it is that the observed values\
could be considered a random draw.

#### Monte Carlo Moran's I

For percent_white

```{r}
moran.mc(tracts_sp$percent_white, ww, nsim=99)
```

For median_income

```{r}
moran.mc(tracts_sp$median_income, ww, nsim=99)
```

#### INTERPRETATION

From the global Moran's I we can conclude that both the percent_whit
and\
median_income are clustered globally. The assumption that the predictor\
percent_white is independent is violated. Both variables show spatial\
auto-correlation.

### Local Moran's I Statistic - Where are the clusters?

Get a labeled set of tracts

```{r}
tracts_sp_named <- data.frame(tracts_sp,row.names=tracts_sp$fips)
```

#### For percent_white

```{r}
lm_pw <- localmoran(tracts_sp$percent_white,ww)
lm_pw_named <- data.frame(lm_pw,row.names=tracts_sp$fips)
lm_pw_sp <- merge(tracts_sp_named, lm_pw_named, by=0) 
percent_white <- summary(lm_pw_sp$percent_white)
zscores <- summary(lm_pw_sp$Z.Ii) # Z-scores for Local Moran's I
```

| Note that the following divisions seem fairly arbitrary. In fact there is
| literature that suggests that p \< 0.05 is not useful with Local Moran's I.

```{r}
lm_pw_sp$sig <- lm_pw_sp$Pr.z....E.Ii.. < 0.05
lm_pw_sp$highv <- lm_pw_sp$percent_white > percent_white['Mean']
lm_pw_sp$lowv <- lm_pw_sp$percent_white < percent_white['Mean']
lm_pw_sp$highz <- lm_pw_sp$Z.Ii > zscores['Mean']
lm_pw_sp$lowz <- lm_pw_sp$Z.Ii < zscores['Mean']
lm_pw_sp$q <- 'ns'
```

Create a categorical column for easy mapping.

```{r}
lm_pw_sp <- within(lm_pw_sp,{
  q[sig & highv & highz]='HH'
  q[sig & lowv & lowz]='LL'
  q[sig & lowv & highz]='LH'
  q[sig & highv & lowz]='HL'
})
```

Show percent_white clusters according to Local Moran's I.

```{r}
options(repr.plot.width = 16, repr.plot.height = 16)
ggplot() + 
  geom_sf(data=lm_pw_sp$geometry,aes(fill = lm_pw_sp$q)) + 
  scale_fill_manual(values=c('HH'="#e31a1c",'LL'="#CCCCFF", 'LH'="#1f78b4", 'HL'="#FFCCCC", 'ns'="#FFFFFF"))
options(repr.plot.width = 8, repr.plot.height = 8)
```

#### INTERPRETATION

HH is where tracts that have percent white are surrounded by tracts with
high percent white - a cluster. Same for LH. These are tracts where low
percent white tracts are surrounded by low percent white tracts. Both
the HL and LL are tracts that have either high or low percent white and
are surrounded by the opposite - the anomalies.

#### For median_income

```{r}
lm_mi <- localmoran(tracts_sp$median_income,ww)
lm_mi_named <- data.frame(lm_mi,row.names=tracts_sp$fips)
lm_mi_sp <- merge(tracts_sp_named, lm_mi_named, by=0) 
median_income <- summary(lm_mi_sp$median_income)
zscores <- summary(lm_mi_sp$Z.Ii) # Z-scores for Local Moran's I
```

Make the bins

```{r}
lm_mi_sp$sig <- lm_mi_sp$Pr.z....E.Ii.. < 0.05
lm_mi_sp$highv <- lm_mi_sp$median_income > median_income['Mean']
lm_mi_sp$lowv <- lm_mi_sp$median_income < median_income['Mean']
lm_mi_sp$highz <- lm_mi_sp$Z.Ii > zscores['Mean']
lm_mi_sp$lowz <- lm_mi_sp$Z.Ii < zscores['Mean']
lm_mi_sp$q <- 'ns'
```

Create a categorical column for easy mapping

```{r}
lm_mi_sp <- within(lm_mi_sp,{
  q[sig & highv & highz]='HH'
  q[sig & lowv & highz]='LH'
  q[sig & lowv & lowz]='LL'
  q[sig & highv & lowz]='HL'
})
```

Show percent_white clusters according to Local Moran's I

```{r}
options(repr.plot.width = 16, repr.plot.height = 16)
ggplot() + 
  geom_sf(data=lm_mi_sp$geometry,aes(fill = lm_mi_sp$q)) + 
  scale_fill_manual(values=c('HH'="#e31a1c",'LL'="#CCCCFF", 'LH'="#1f78b4", 'HL'="#FFCCCC", 'ns'="#FFFFFF"))
options(repr.plot.width = 8, repr.plot.height = 8)
```

#### INTERPRETATION

HH is where tracts that have high income are surrounded by tracts with
high income - a cluster. Same for LH. These are tracts where low income
tracts are surrounded by low income tracts. Both the HL and LL are
tracts that have either high or low median income surrounded by the
opposite with no clustering - the anomalies.

Take a moment to look back and for between the maps. Do the clusters
appear in the same location? Would your observations change the way you
interpret the basic linear model?

### Spatial Lag Regression

Spatial lag regression provides an alternative model that takes into
account the auto-correlation.

$y = \rho W y + X \beta + \varepsilon$

where $\rho$ is found by optimize() first, and $\beta$ and other\
parameters by generalized least squares subsequently (one-dimensional\
search using optim performs badly on some platforms). In the spatial\
Durbin (mixed) model, the spatially lagged independent variables are
added to X.

```{r}
lagged <- lagsarlm(
  percent_white~median_income_of_residents,
  data=tracts_sp, 
  listw=ww, 
  tol.solve=1.0e-30, 
  zero.policy=T)
summary(lagged)
```

#### INTERPRETATION

The lower AIC value for the spatial lag model relative to the\
non-spatial model indicates that the spatial lag model has a slightly\
better fit than the non-spatial model.

Make a residuals plot

```{r}
x = data.frame(actual = lagged$fitted.values + lagged$residuals, fitted = lagged$fitted.values)
x = x[order(x$fitted),]
plot(x$actual, col='#00000040')
lines(x$fitted, col='#c00000', lwd=3)
```

#### INTERPRETATION

The lagged model and the linear model are actually still pretty close,
***BUT*** how does the cluster analysis influence how you interpret the
model? It seems that there are other important factors to predict median
income other than whiteness that we are not considering .... more work
to do.

### Geographically Weighted Regression

***NOTE: This package does not constitute approval of GWR\
as a method of spatial analysis; see example(gwr)***

In other words, everything below this point is experimental and not
really\
working right or to be trusted if it actually runs ....

Spatially Weighted Regression is not a true regression (i.e. predictive\
model), but instead tries to predict where

> non-stationarity is taking place on the map, that is where locally\
> weighted regression coefficients move away from their global values.

see this vignette:\
<https://cran.r-project.org/web/packages/spgwr/vignettes/GWR.html#Geographically_Weighted_Regression_1>
